\documentclass[12pt]{strath_thesis}

\usepackage{apacite}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[mathscr]{eucal}
\usepackage{ifpdf}
\usepackage{multirow}
\usepackage{dcolumn}
\usepackage{eurosym}
\usepackage{rotating}
\usepackage{pdflscape}
\usepackage{paralist}
\usepackage{ifthen}

%\usepackage{pslatex}
%\usepackage{palatino}

% \ifpdf
%   \usepackage[pdftex]{graphicx}
%   \graphicspath{{./pdf/}{./jpg/}}
%   \DeclareGraphicsExtensions{.pdf,.jpg,.png}
% \else
%   \usepackage[dvips]{graphicx}
%   \graphicspath{{./eps/}}
%   \DeclareGraphicsExtensions{.eps}
% \fi

\usepackage[ruled]{algorithm}
\usepackage{algorithmic}

\usepackage{ulem}

\usepackage{tikz}
\usetikzlibrary{arrows,calc}
\usetikzlibrary{shadows}
\usetikzlibrary{decorations,decorations.pathmorphing}
\usetikzlibrary{positioning,shapes}
\tikzset{
>=stealth', % standard arrow tip
help lines/.style={dashed, thick}, % different line styles
axis/.style={<->},
important line/.style={thick},
connection/.style={thick, dotted},
}

% PGF/TikZ sequence diagrams.
\usepackage[underline=true,rounded corners=false]{pgf-umlsd}

\input{tikz/tikz_cmd}

\newcolumntype{d}[1]{D{.}{\cdot}{#1}}
\newcolumntype{.}[1]{D{.}{.}{#1}}

\makeatletter
\newcommand{\rmnum}[1]{\romannumeral #1}
\newcommand{\Rmnum}[1]{\expandafter\@slowromancap\romannumeral #1@}
\makeatother
\newcommand{\abs}[1]{\lvert#1\rvert}
\newcommand{\diag}{\mathop{\mathbf{diag}}}
\newcommand{\matpower}{\textsc{Matpower~}}
\newcommand{\matlab}{Matlab~}
\newcommand{\psat}{\textsc{Psat~}}
\newcommand{\pylon}{Pylon~}
\newcommand{\stable}{$\bullet$}
\newcommand{\unstable}{$\bullet$}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}}

% \newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\tiny\bf
% +}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\tiny\bf +}}
\def\CC{{C\nolinebreak[4]\hspace{-.05em}\raisebox{.4ex}{\tiny\bf ++}}}

% Optional argument for the parenthesis in bordermatrix.
\makeatletter
\newif\if@borderstar
\def\bordermatrix{\@ifnextchar*{%
\@borderstartrue\@bordermatrix@i}{\@borderstarfalse\@bordermatrix@i*}%
}
\def\@bordermatrix@i*{\@ifnextchar[{\@bordermatrix@ii}{\@bordermatrix@ii[()]}}
\def\@bordermatrix@ii[#1]#2{%
\begingroup
\m@th\@tempdima8.75\p@\setbox\z@\vbox{%
\def\cr{\crcr\noalign{\kern 2\p@\global\let\cr\endline }}%
\ialign {$##$\hfil\kern 2\p@\kern\@tempdima & \thinspace %
\hfil $##$\hfil && \quad\hfil $##$\hfil\crcr\omit\strut %
\hfil\crcr\noalign{\kern -\baselineskip}#2\crcr\omit %
\strut\cr}}%
\setbox\tw@\vbox{\unvcopy\z@\global\setbox\@ne\lastbox}%
\setbox\tw@\hbox{\unhbox\@ne\unskip\global\setbox\@ne\lastbox}%
\setbox\tw@\hbox{%
$\kern\wd\@ne\kern -\@tempdima\left\@firstoftwo#1%
\if@borderstar\kern2pt\else\kern -\wd\@ne\fi%
\global\setbox\@ne\vbox{\box\@ne\if@borderstar\else\kern 2\p@\fi}%
\vcenter{\if@borderstar\else\kern -\ht\@ne\fi%
\unvbox\z@\kern-\if@borderstar2\fi\baselineskip}%
\if@borderstar\kern-2\@tempdima\kern2\p@\else\,\fi\right\@secondoftwo#1 $%
}\null \;\vbox{\kern\ht\@ne\box\tw@}%
\endgroup
}
\makeatother

% Remove figures to speed up build (should be possible using 'draft').
\newboolean{includefigures}
\setboolean{includefigures}{false}
% Select chapters to build.
\includeonly{
%	introduction,
%  	background,
%  	related_work,
%  	method,
   	results,
   	conclusion,
   	appendix
}

\ifpdf
\pdfinfo{/Author (Richard W. Lincoln)
  		 /Title  (Learning to Trade Power)
         /Subject (Electrical Power Engineering)}
\fi

\title{Learning to Trade Power}
%\title{Learning Power Trade}
\author{Richard W. Lincoln}

\pagestyle{plain}

\begin{document}

\maketitle

\setcounter{page}{1}
\pagenumbering{roman}

\declaration

\begin{acknowledgements}
I wish to thank Professor Jim McDonald for giving me the opportunity to study
at The Institute for Energy and Environment and for giving me the freedom
to pursue my own research interests.  I also wish to thank my supervisors,
Professor~Graeme~Burt and Dr~Stuart~Galloway, for their guidance and
 scholarship.  I wish to offer very special thanks to my parents, my big
brother and my little sister for all of their support throughout my PhD.

This thesis makes extensive use of open source software projects developed by
researchers from other institutions.  I wish to thank Dr Ray Zimmerman from
Cornell University for his work on optimal power flow, researchers from the Dalle Molle Institute for
Artificial Intelligence (IDSIA) and the Technical University of Munich for
their work on reinforcement learning algorithms and artificial neural networks
and Charles Gieseler from Iowa State University for his implementation of the
Roth-Erev reinforcement learning method.

This research was funded by the United Kingdom Engineering and Physical
Sciences Research Council through the Supergen Highly Distributed Power
Systems consortium under grant GR/T28836/01.
\end{acknowledgements}

\begin{abstract}
In Electrical Power Engineering, learning algorithms can be used to model
the strategies of electricity market participants.  The objective of this work
is to establish if \textit{policy gradient} reinforcement learning algorithms
can be used to create participant models superior to those using previously
applied \textit{value function} based methods.

% Electricity is an unique commodity.  It must be supplied at the presise moment
% that it is demanded, as there are no practical methods for significant levels
% of storage in most systems.  There are also few options for controlling the
% direction in which power flows around networks, so generation or consumption
% at one point in the system has an effect at all other points.
Supply of electricity involves technology, money, people, natural resources and
the environment.  All of these aspects are changing and electricity market
designs must be suitably researched to ensure that they are fit for purpose.  In
this thesis electricity markets are modelled as non-linear constrained
optimisation problems, which are solved using a primal-dual interior point
method.  Policy gradient reinforcement learning algorithms are used to adjust
the parameters of multi-layer feed-forward artificial neural networks that
approximate each market participant's policy for selecting power quantities and
prices that are offered in the simulated marketplace.

Traditional reinforcement learning methods that learn a value function have been
previously applied in simulated electricity trade, but are largely restricted to
discrete representations of a market environment. Policy gradient methods have
been proven to offer convergence guarantees in continuous environments and avoid
many of the problems that mar value function based methods.

% The benefits of using policy gradient methods in electricity market simulation
% are explored and the results demonstrate their superior trading ability when
% operating in large constrained networks.
% By introducing advanced unsupervised learning methods to electricity market
% simulation, previous research in the field may be revisited and enhanced and
% an opportunity to use policy gradient methods in decision support and automated
% energy trading applications is opened.

% Reinforcement learning methods that use connectionist systems for value
% function approximation offer few convergence guarantees, even in simple
% systems.  Table-based value function reinforcement learning methods have been
% used previously for the simulation of electricity markets, but they operate
% only in discrete action and sensor domains.  If learning algorithms are to
% deliver on their potential for application in operational settings then it will
% be necessary for them to operate in continuous domains.  The principle
% contribution of this thesis is the demonstration of policy-gradient
% reinforcement learning algorithms being applied to continuous representations
% of electricity trading problems, showing that superior use of sensor data
% results in improved overall performance when compared with previously applied
% value-function methods.  From this it follows that learning methods which
% search directly in the policy space will be better suited to decision support
% applications and automated electric power trade.
\end{abstract}

\tableofcontents
\newpage

\listoffigures
\addcontentsline{toc}{chapter}{List of Figures}
\newpage

\listoftables
\addcontentsline{toc}{chapter}{List of Tables}
\newpage

\pagenumbering{arabic}
\setcounter{page}{1}

\onehalfspacing

% Set the scene and problem statement. Introduce structure of thesis, state
% contributions (3-5).
\include{introduction}

% Demonstrate wider appreciation (context). Provide motivation.
\include{background}

% Survey and critical assessment. Relation to own work.
\include{related_work}

% Analysis, design, implementation and interpretation of results.
\include{method}

% Critical assessment of own work.
\include{results}

% State hypothesis. Further Work. Restate contribution.
\include{conclusion}

\bibliographystyle{apacite}
\renewcommand{\bibname}{Bibliography}
\bibliography{literature}
%\addcontentsline{toc}{chapter}{Bibliography}

\appendix
\include{appendix}

\end{document}
